---
title: "propionate_hmmer_scorefilter"
output: html_document
---
# purpose of this doc: assess hmmer score distribution for known propionate producers that use the Wood-Werkman cycle (sub-pathway of P1 propionate pathway)

# find lowest and highest scores for each gene 
```{r}
library(dplyr)

model_P1_strains <- c("Acidipropionibacterium_acidipropionici",
              "Propionibacterium_freudenreichii",
              "Bacteroides_fragilis",
              "Prevotella_ruminocola",
              "Propiogenium_modestum",
              "Selenomonas_ruminantium",
              "Selenomonas_sputigena",
              "Veillonella_parvula")

model_P1_df <- (subset(maxHitScoreDF_P1, 
                       grepl(model_P1_strains[1], strain) | 
                          grepl(model_P1_strains[2], strain) |
                         grepl(model_P1_strains[3], strain) | 
                         grepl(model_P1_strains[4], strain) | 
                         grepl(model_P1_strains[5], strain) | 
                         grepl(model_P1_strains[6], strain) | 
                         grepl(model_P1_strains[7], strain) | 
                         grepl(model_P1_strains[8], strain) 
                       == TRUE))

model_WWC_df <- (subset(maxHitScoreDF_P1, 
                       grepl(model_P1_strains[1], strain) | 
                          grepl(model_P1_strains[2], strain)
                       == TRUE))

model_succ_df <- (subset(maxHitScoreDF_P1, 
                         grepl(model_P1_strains[3], strain) | 
                         grepl(model_P1_strains[4], strain) | 
                         grepl(model_P1_strains[5], strain) | 
                         grepl(model_P1_strains[6], strain) | 
                         grepl(model_P1_strains[7], strain) | 
                         grepl(model_P1_strains[8], strain)
                       == TRUE))

# lowest model score per gene
# mutase and epimerase
slice(subset(model_P1_df, model_P1_df$gene == "methylmalonyl-CoA-mutase-EC5-4-99-2"), which.min(hit_score))

slice(subset(model_P1_df, model_P1_df$gene == "methylmalonyl-CoA-epimerase-EC5-1-99-1"), which.min(hit_score))

# carboxyltransferase subunits
slice(subset(model_WWC_df, model_WWC_df$gene == "methylmalonyl-CoA-carboxyltransferase12S-KOK17489"), which.min(hit_score))
slice(subset(model_WWC_df, model_WWC_df$gene == "methylmalonyl-CoA-carboxyltransferase5S-KOK03416"), which.min(hit_score))
slice(subset(model_WWC_df, model_WWC_df$gene == "methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490"), which.min(hit_score))

# decarboxylase subunits
slice(subset(model_succ_df, model_succ_df$gene == "methylmalonyl-CoA-decarboxylaseAlpha-KOK01604"), which.min(hit_score))
slice(subset(model_succ_df, model_succ_df$gene == "methylmalonyl-CoA-decarboxylaseBeta-KOK20509"), which.min(hit_score))
slice(subset(model_succ_df, model_succ_df$gene == "methylmalonyl-CoA-decarboxylaseGamma-KOK23351"), which.min(hit_score))
slice(subset(model_succ_df, model_succ_df$gene == "methylmalonyl-CoA-decarboxylaseDelta-KOK23352"), which.min(hit_score))

# mininum model hit scores per gene
mutase_min_score = 768.3
epimerase_min_score = 175.5
carboxyltransferase12S_min_score = 904.5
carboxyltransferase5S_min_score = 1003.1
carboxyltransferase1_3S_min_score = 171.9
decarboxylaseAlpha_min_score = 889.1
decarboxylaseBeta_min_score = 558.1
decarboxylaseGamma_min_score = 146.4
decarboxylaseDelta_min_score = 98.1

```

# let's look at the results of the HMMER search for all strains
# specifically, the distribution of hit_scores for each gene
```{r}
library(ggplot2)

filter_factor = 0.5
filter_factor2 = 0.75

# score distribution for mutase and epimerase
ggplot(subset(maxHitScoreDF_P1, gene %in% c("methylmalonyl-CoA-epimerase-EC5-1-99-1", "methylmalonyl-CoA-mutase-EC5-4-99-2")), aes(x = hit_score)) +
  geom_histogram(binwidth = 10) +
  facet_wrap(vars(gene)) +
  
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-epimerase-EC5-1-99-1"), aes(xintercept = epimerase_min_score *filter_factor), color = "red") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-mutase-EC5-4-99-2"), aes(xintercept = mutase_min_score *filter_factor), color = "red") +
  
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-epimerase-EC5-1-99-1"), aes(xintercept = epimerase_min_score *filter_factor2), color = "blue") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-mutase-EC5-4-99-2"), aes(xintercept = mutase_min_score *filter_factor2), color = "blue")


# distribution of scores for carboxyltransferase subunits
ggplot(subset(maxHitScoreDF_P1, gene %in% c("methylmalonyl-CoA-carboxyltransferase5S-KOK03416", "methylmalonyl-CoA-carboxyltransferase12S-KOK17489", "methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490")), aes(x = hit_score)) +
  geom_histogram(binwidth = 10) +
  facet_wrap(vars(gene)) +
  
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-carboxyltransferase5S-KOK03416"), aes(xintercept = carboxyltransferase5S_min_score *filter_factor), color = "red") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-carboxyltransferase12S-KOK17489"), aes(xintercept = carboxyltransferase12S_min_score *filter_factor), color = "red") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490"), aes(xintercept = carboxyltransferase1_3S_min_score *filter_factor), color = "red") +
  
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-carboxyltransferase5S-KOK03416"), aes(xintercept = carboxyltransferase5S_min_score *filter_factor2), color = "blue") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-carboxyltransferase12S-KOK17489"), aes(xintercept = carboxyltransferase12S_min_score *filter_factor2), color = "blue") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490"), aes(xintercept = carboxyltransferase1_3S_min_score *filter_factor2), color = "blue")

# distribution of scores for decarboxylase subunits
ggplot(subset(maxHitScoreDF_P1, gene %in% c("methylmalonyl-CoA-decarboxylaseAlpha-KOK01604", "methylmalonyl-CoA-decarboxylaseBeta-KOK20509", "methylmalonyl-CoA-decarboxylaseGamma-KOK23351", "methylmalonyl-CoA-decarboxylaseDelta-KOK23352")), aes(x = hit_score)) +
  geom_histogram(binwidth = 10) +
  facet_wrap(vars(gene)) +
  
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseAlpha-KOK01604"), aes(xintercept = decarboxylaseAlpha_min_score *filter_factor), color = "red") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseBeta-KOK20509"), aes(xintercept = decarboxylaseBeta_min_score *filter_factor), color = "red") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseGamma-KOK23351"), aes(xintercept = decarboxylaseGamma_min_score *filter_factor), color = "red") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseDelta-KOK23352"), aes(xintercept = decarboxylaseDelta_min_score *filter_factor), color = "red") +
  
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseAlpha-KOK01604"), aes(xintercept = decarboxylaseAlpha_min_score *filter_factor2), color = "blue") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseBeta-KOK20509"), aes(xintercept = decarboxylaseBeta_min_score *filter_factor2), color = "blue") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseGamma-KOK23351"), aes(xintercept = decarboxylaseGamma_min_score *filter_factor2), color = "blue") +
  geom_vline(data=filter(maxHitScoreDF_P1, gene == "methylmalonyl-CoA-decarboxylaseDelta-KOK23352"), aes(xintercept = decarboxylaseDelta_min_score *filter_factor2), color = "blue")
  

```

# subset maxHitScoreDF for score filter of 75% of lowest scoring model propionate producer
# FIRST FILTERING METHOD: SCORE CUTOFF
```{r}
filter_factor = 0.75
P1_score_filtered_df <- as.data.frame(maxHitScoreDF_P1) %>%
  filter((gene == "methylmalonyl-CoA-epimerase-EC5-1-99-1" & hit_score >= epimerase_min_score * filter_factor) |
         (gene == "methylmalonyl-CoA-mutase-EC5-4-99-2" & hit_score >= mutase_min_score * filter_factor) |
         (gene == "methylmalonyl-CoA-carboxyltransferase5S-KOK03416" & hit_score >= carboxyltransferase5S_min_score * filter_factor) |
         (gene == "methylmalonyl-CoA-carboxyltransferase12S-KOK17489" & hit_score >= carboxyltransferase12S_min_score * filter_factor) |
         (gene == "methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490" & hit_score >= carboxyltransferase1_3S_min_score * filter_factor) |
           (gene == "methylmalonyl-CoA-decarboxylaseAlpha-KOK01604" & hit_score >= decarboxylaseAlpha_min_score * filter_factor) |
           (gene == "methylmalonyl-CoA-decarboxylaseBeta-KOK20509" & hit_score >= decarboxylaseBeta_min_score * filter_factor) |
           (gene == "methylmalonyl-CoA-decarboxylaseGamma-KOK23351" & hit_score >= decarboxylaseGamma_min_score * filter_factor) |
           (gene == "methylmalonyl-CoA-decarboxylaseDelta-KOK23352" & hit_score >= decarboxylaseDelta_min_score * filter_factor))

# reshape filtered_df from long to wide by gene (gene)

P1_score_filtered_wide <- reshape(P1_score_filtered_df, timevar = "gene", direction = "wide", idvar = "strain")
View(P1_score_filtered_wide)

write.csv(filtered_wide, "0.75scoreFilter_P1_HMMER_hits.csv")
```

# subset score filtered df by number of genes present
# SECOND FILTERING METHOD: NUMBER OF GENES
```{r}
# requirements for genomes:
# must have mutase, epimerase, and either ALL carboxyltransferase subunits or ALL decarboxylase subunits
# (include decarboxylase delta subunit for first iteration. can remove as a requirement later on)
# FIRST: run HMMER search on model genomes to determine appropriate filter for # of gene requirement
# done. must have ALL subunits for either the carboxyltransferase or the decarboxylase based on HMMER hits in model strains.

P1_score_ngenes_filtered_wide <- subset(P1_score_filtered_wide, 
                                        !is.na(`hit_score.methylmalonyl-CoA-epimerase-EC5-1-99-1` &
                                                 `hit_score.methylmalonyl-CoA-mutase-EC5-4-99-2`) & 
                                          (!is.na(`hit_score.methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490` & 
                                                            `hit_score.methylmalonyl-CoA-carboxyltransferase12S-KOK17489` &
                                                            `hit_score.methylmalonyl-CoA-carboxyltransferase5S-KOK03416`) | 
                                                     !is.na(`hit_score.methylmalonyl-CoA-decarboxylaseAlpha-KOK01604` & 
                                                              `hit_score.methylmalonyl-CoA-decarboxylaseBeta-KOK20509` &
                                                              `hit_score.methylmalonyl-CoA-decarboxylaseDelta-KOK23352` &
                                                              `hit_score.methylmalonyl-CoA-decarboxylaseGamma-KOK23351`))
)

View(P1_score_ngenes_filtered_wide)

# add column to distinguish which P1 pathway

P1_score_ngenes_filtered_wide$pathway <- ifelse(!is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-carboxyltransferase1-3S-KOK17490`) & 
                                          !is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-carboxyltransferase12S-KOK17489`) &
                                          !is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-carboxyltransferase5S-KOK03416`),
                                          "Wood-Werkman Cycle",
                                          ifelse(!is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-decarboxylaseAlpha-KOK01604`) & 
                                          !is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-decarboxylaseBeta-KOK20509`) &
                                          !is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-decarboxylaseDelta-KOK23352`) &
                                          !is.na(P1_score_ngenes_filtered_wide$`hit_score.methylmalonyl-CoA-decarboxylaseGamma-KOK23351`), "sodium-pumping decarboxylase", NA))

write.csv(P1_score_ngenes_filtered_wide, "0.75scoreFilter_nGenes_P1_HMMER_hits.csv")

```


